---
title: "Minnesota: Nitrate in Drinking Water"
author: "Joe Wertz"
date: "10/22/2019"
output: html_document

---

# About this analysis  

Analysis of drinking water testing data to find out how many people were customers of municipal water systems in Minnesota that tested over EPA limit of 10 mg/l for nitrate.  

## Data source:
Minnesota Department of Health drinking water query: https://data.web.health.state.mn.us/web/mndata/drinkingwater_query

- selected all counties, all dates for contaminant "nitrate" 
- downloaded .csv

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include = FALSE}
## Import libraries
library(tidyr)
library(readr)
library(skimr)
library(dplyr)
library(stringr)
```

### Import the data, set pwsId as text instead of numeric
```{r import}
## import .csv from MNSDWIS
water <- read_csv("water.csv")
cols(
  pwsId = col_character(),
  year = col_date()
  )
```

### Drop the empty "x12" column created with the extra comma in the data
```{r drop_column}
water <- water %>% select(-X12)
```

### Create new number-only columns for mean and max
```{r new_columns}
water$mean_num <- gsub("[^0-9\\.]","",water$mean)
water$max_num <- gsub("[^0-9\\.]","",water$max)
# convert new column to numeric
water$mean_num <- as.numeric(water$mean_num)
water$max_num <- as.numeric(water$max_num)

```

### Drop rows 18184-18194, which include non-data like source info, url, etc.
```{r drop_rows}
water <- water %>% filter(row_number() <18183)
```

### Filter for water systems that have mean > 10 mg/l EPA limit
```{r, over_limit_mean}
over_limit_mean <- water %>%  filter(mean_num >= 10)
```

### filter for water systems that have max > 10 mg/l EPA limit and summarize to get total population potentially exposed
```{r over_limit_max}
over_limit_max <- water %>%  filter(max_num >= 10)
over_limit_max %>% 
 summarise(total_pop = sum(populationServed))
```


